---
title: "R Notebook"
output: html_notebook
---
Missing Data
In this assignment you will complete a variety of tasks related to working with missing data.

Libraries: For this assignment you will need the following libraries: tidyverse, VIM, and mice.

```{r libraries}

library(tidyverse)
library(VIM)
library(mice)
library(skimr)

```

Before beginning the assignment tasks, you should read-in the data for the assignment into a data frame
called grades. This data contains grade information from an engineering course. 

The dataset was originally used to investigate how student performance in the course would be predictive of student grades on the
“Final” exam. The “Prefix” column is a surrogate for enrollment year in the engineering program. Smaller
values imply older (more mature?) students.

```{r data load}

grades <- read_csv("class-grades.csv")

```


Task 1: For which variables is there missing data?
```{r T1}

skim(grades)

```


Task 2: Use the VIM package to visualize missingness. Are there any students that are missing multiple
pieces of data?

```{r T2}

vim_plot = aggr(grades, numbers = TRUE, prop = c(TRUE, FALSE),cex.axis=.7)

```

Task 3: Use row-wise deletion of missing values to create a new data frame. How many rows remain in
this data frame?

```{r T3}

grades_rowdel <- grades |> drop_na()

skim(grades_rowdel)

```

Task 4: Use column-wise deletion of missing values to create a new data frame (from the original data
frame not from the data frame created in Task 3). How many columns remain in this data frame?


```{r T4}
grades_coldel <- grades |> select(-Tutorial,-Midterm,-TakeHome,-Final)
```

Task 5: Which approach (Task 3 or Task 4) seems preferable for this dataset? Briefly discuss your answer.

  Deleting rows with missing data seems much more preferable for this data set. We lose only about 10% of the rows but about 67% of the columns when we delete the NAs. The remaining data from the row deletion (89 observations of 6 variables) will still be meaningful for our analysis.
  
  
Task 6 Use the “mice” package to impute the missing data in this dataset. Set “m = 5” and “method
=”pmm"" in the imputation. You MUST use a random number seed of 123 for this imputation. What is the
mean value for the “Final” variable after imputation?

```{r T5}

set.seed(123)
imp_grades = mice(grades, m=5, method = "pmm", printFlag = FALSE)

summary(imp_grades)

```


```{r T5b}

densityplot(imp_grades,~Final)
```
```{r T5c}

grades_complete = complete(imp_grades)
summary(grades_complete)

```

