---
title: "Mod4.Assign3"
author: "Hilary Ross"
date: "2026-06-06"
output:
  word_document: default
  html_document: default
---

## Random Forests

Libraries: For this assignment you will need the following libraries: tidyverse, tidymodels, caret, gridExtra,
vip, and ranger.

```{r Libraries}

library(tidyverse)
library(tidymodels)
library(caret)
library(gridExtra)
library(vip)
library(ranger)
library(skimr)


```

Read in the “drug_data.csv” dataset. This dataset deals with drug consumption in individuals across a wide
spectrum of countries, drugs, ages, etc. A description of the dataset is available here: http://archive.ics.uci.
edu/ml/datasets/Drug+consumption+%28quantified%29

```{r Data Load}

drug <- read_csv("drug_data-2.csv")

```


This dataset (as is common with medical and healthcare datasets) requires quite a bit of cleaning and
preparation before analysis.

I’ll walk you through the cleaning before we start our random forest work.

Loading the Data The columns do not have names, so we’ll supply names via the names function.

```{r Names}

names(drug) = c("ID", "Age", "Gender", "Education", "Country", "Ethnicity",
"Nscore", "Escore", "Oscore", "Ascore", "Cscore", "Impulsive",
"SS", "Alcohol", "Amphet", "Amyl", "Benzos", "Caff", "Cannabis",
"Choc", "Coke", "Crack", "Ecstasy", "Heroin", "Ketamine", "Legalh",
"LSD", "Meth", "Mushrooms", "Nicotine", "Semer", "VSA")

```

Next-up we change all CL0 and CL1 values to “No” and CL2, CL3, CL4, CL5, and CL6 values to “Yes”.
CL0 and CL1 imply the drug was never used or used over a decade ago. CL2 through CL6 imply more
recent drug use. The code below finds any CL0, CL1, etc. values in the data frame and replaces them with
the appropriate “No” or “Yes”.

```{r Update Values}

drug[drug == "CL0"] = "No"
drug[drug == "CL1"] = "No"
drug[drug == "CL2"] = "Yes"
drug[drug == "CL3"] = "Yes"
drug[drug == "CL4"] = "Yes"
drug[drug == "CL5"] = "Yes"
drug[drug == "CL6"] = "Yes"

```


Next up we do a good bit of factor conversion and recoding. Note the use of mutate_at to target specific
ranges of variables. You may see a warning message about the use of the funs() function. It is OK to ignore
this.

```{r Factor Conversion and Recoding}

drug_clean = drug %>% mutate_at(vars(Age:Ethnicity), funs(as_factor)) %>%
  mutate(Age = factor(Age, labels = c("18_24", "25_34", "35_44", "45_54", 
                                      "55_64", "65_"))) %>%
  mutate(Gender = factor(Gender, labels = c("Male", "Female"))) %>%
  mutate(Education = factor(Education, labels = c("Under16", "At16", "At17", "At18",
                                                  "SomeCollege","ProfessionalCert",
                                                  "Bachelors", "Masters",
                                                  "Doctorate"))) %>%
  mutate(Country = factor(Country, labels = c("USA", "NewZealand", "Other", "Australia",
                                              "Ireland","Canada","UK"))) %>%
  mutate(Ethnicity = factor(Ethnicity, labels = c("Black", "Asian", "White",
                                                  "White/Black", "Other",
                                                  "White/Asian", "Black/Asian"))) %>%
  mutate_at(vars(Alcohol:VSA), funs(as_factor)) %>%
  select(-ID)

```

Take a peek at the cleaned data to make sure we are all good before proceeding.

```{r}

str(drug_clean)

```

We’ll focus on Nicotine use, so let’s get rid of the remaining drug use variables. We’ll use select for this.

```{r Select Variables}

drug_clean = drug_clean %>% select(!(Alcohol:Mushrooms)) %>% select(!(Semer:VSA))

```

Now we’re in business. Let’s get started with the tasks.

Question 1: Check for missing data in our “drug_clean” dataframe.

```{r Q1}

skim(drug_clean)

```

True/**False**: There is missingness in the dataset.

Question 2: Split the dataset into training (70%) and testing (30%) sets. Use a set.seed of 1234. Stratify
by the “Nicotine” variable.
How many rows are in the training set?

```{r Q2}

set.seed(1234)
drug_split = initial_split(drug_clean, prop = 0.70, strata = Nicotine)
drug_train = training(drug_split)
drug_test = testing(drug_split)

nrow(drug_train)

```

Question 3: Create appropriate visualizations (12 in all) to examine the relationships between each variable
and “Nicotine”. Use grid.arrange (from the gridExtra package) to organize these visuals (perhaps in groups
of four visualizations?).
**True**/False: Individuals in the 18-24 age group are proportionally more likely to be Nicotine users than not.

```{r Q3}

p1 = ggplot(drug_train, aes(x = Age, fill = Nicotine)) + geom_bar(position = "fill")
p2 = ggplot(drug_train, aes(x = Gender, fill = Nicotine)) + geom_bar(position = "fill")
p3 = ggplot(drug_train, aes(x = Education, fill = Nicotine)) + geom_bar(position = "fill")
p4 = ggplot(drug_train, aes(x = Country, fill = Nicotine)) + geom_bar(position = "fill")
p5 = ggplot(drug_train, aes(x = Ethnicity, fill = Nicotine)) + geom_bar(position = "fill")
p6 = ggplot(drug_train, aes(x = Nscore, fill = Nicotine)) + geom_bar(position = "fill")
p7 = ggplot(drug_train, aes(x = Escore, fill = Nicotine)) + geom_bar(position = "fill")
p8 = ggplot(drug_train, aes(x = Oscore, fill = Nicotine)) + geom_bar(position = "fill")
p9 = ggplot(drug_train, aes(x = Ascore, fill = Nicotine)) + geom_bar(position = "fill")
p10 = ggplot(drug_train, aes(x = Cscore, fill = Nicotine)) + geom_bar(position = "fill")
p11 = ggplot(drug_train, aes(x = Impulsive, fill = Nicotine)) + geom_bar(position = "fill")
p12 = ggplot(drug_train, aes(x = SS, fill = Nicotine)) + geom_bar(position = "fill")

grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,p12)

```


Question 4: **True**/False: Individuals with higher “Impulsive” scores more likely to be Nicotine users than
not.

Question 5: Create a random forest model (using the ranger package) on the training set to predict Nicotine
using all of the variables in the dataset. You 5-fold, k-fold cross-validation (random number seed of 123 for
the folds). Allow R to select mtry values between 2 and 8 and min_n values between 5 and 20. Use 10
levels in your “grid_regular” function. Set a random number seed of 123 for the tune_grid function. Use
100 trees.
NOTE: This model may take a few minutes to run. Be patient :)
Visualize the relationships between parameters and performance metrics.
The highest accuracy in this visualization is just greater than which value:
A. 0.725
**B. 0.730**
C. 0.720
D. 0.715

```{r Q5}

set.seed(123)
rf_folds = vfold_cv(drug_train, v = 5)

drug_recipe = recipe(Nicotine ~., drug_train) %>%
  step_dummy(all_nominal(), -all_outcomes())

rf_model = rand_forest(mtry = tune(), min_n = tune(), trees = 100) %>% 
  set_engine("ranger",importance = "permutation") %>% 
  set_mode("classification")

drug_wflow = 
  workflow() %>% 
  add_model(rf_model) %>% 
  add_recipe(drug_recipe)

rf_grid = grid_regular(
  mtry(range = c(2, 8)), #these values determined through trial and error (or provided by instructor)
  min_n(range = c(5, 20)), #these values determined through significant trial and error (or provided by instructor)
  levels = 10
)

set.seed(123)
rf_res = tune_grid(
  drug_wflow,
  resamples = rf_folds,
  grid = rf_grid #use the tuning grid
)

rf_res %>%
  collect_metrics() %>%
  filter(.metric == "accuracy") %>%
  select(mean, min_n, mtry) %>%
  pivot_longer(min_n:mtry,
    values_to = "value",
    names_to = "parameter"
  ) %>%
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "Accuracy")
```


Question 6: Use the best mtry and min_n values from Question 5 to finalize the workflow and fit the model
to training set. Examine variable importance.
Which variable is most important?
A. Oscore
B. Cscore
C. Impulsive
**D. SS**

```{r Q6}

best_rf = select_best(rf_res, metric = "accuracy")

final_rf = finalize_workflow(
  drug_wflow,
  best_rf
)

final_rf_fit = fit(final_rf, drug_train)

final_rf_fit %>% pull_workflow_fit() %>% vip(geom = "point")

```

Question 7: To four decimal places, what is the accuracy of your model on the training set?

**Note for the correct answer, balanced accuracy of 0.9143 needed to be used**

```{r Q7}

trainpredrf = predict(final_rf_fit, drug_train)

confusionMatrix(trainpredrf$.pred_class, drug_train$Nicotine, 
                positive = "Yes")

```

Question 8: To four decimal places, what is the naive accuracy (training set)?

```{r}

table(drug_train$Nicotine) %>% prop.table()

```

Question 9: To four decimal places, what is your model’s accuracy on the testing set?

```{r Q9}

drug_test_pred = predict(final_rf_fit, drug_test, type = "class") |> 
  bind_cols(drug_test |> select(Nicotine))

conf_mat_test = confusionMatrix(drug_test_pred$.pred_class,
                           drug_test_pred$Nicotine,
                           positive="Yes")

round(conf_mat_test$overall["Accuracy"],4)


```

Question 10 The difference in accuracy between the training and testing sets implies?
A. Overfitting does not appear to be occurring
**B. Overfitting is likely occurring**
C. It is not clear whether or not overfitting is occurring